05 - MCMCglmm analyses

### 2021-04-05 Analysis/05-mcmcglmm_analyses.Rmd
Update 05-mcmcglmm_analyses.Rmd

```{r}
print("ha")
```

>>>>

```{r, eval = TRUE, print = TRUE, echo = FALSE}
library(MCMCglmm)

## Add the animal and clade (which is more like a clade + a grade) column to data
mcmcglmm_data <- as.data.frame(pca$x)
mcmcglmm_data <- cbind(mcmcglmm_data,
                       clade = as.factor(c(rep("clade", 5), rep("grade", 4))),
                       animal = rownames(pca$x))
## Selecting the traits that have a non zero variance (0 =< 1e-6: for some reason that's what MCMCglmm likes: TODO:fix: see bugs with Choleski decomposition and rounding estimates errors in MCMCglmm)
select_traits <- apply(pca$x, 2, var) > 2e-5

## TEST WITH ONLY THREE TRAITS
select_traits <- c(1,1)

## MCMC params
MCMCparams <- list(nitt   = 110000,
                   burnin = 10000,
                   thin   = 100)

## Making the fixed terms formula
#TODO: do some wrapping to get all the variable names automatically
formula_noclade = cbind(Comp1, Comp2) ~ trait-1
formula_clade = cbind(Comp1, Comp2) ~ trait:clade-1

## Base model
#GT: The simplest model simply estimates the overal covariance matrix for the whole data set
## The base model residuals
base_residuals <- ~ us(trait):units
## Running the base model
model_base <- MCMCglmm(fixed  = formula_noclade,
                       family = rep("gaussian", sum(select_traits)),
                       rcov   = base_residuals,
                       data   = mcmcglmm_data,
                       nitt   = MCMCparams$nitt,
                       burnin = MCMCparams$burnin,
                       thin   = MCMCparams$thin)

## Clade model
#GT: The next model, still without phylogeny, involves modelling clade effect. Note that we are using clade both in the fixed effects and in the random effects. The random effects relate to covariances and main effects to means
## The residuals for the clade model
#TODO: generate automatically from the levels of the data
clade_residuals <- ~ us(at.level(clade,1):trait):units + us(at.level(clade,2):trait):units
## The clade model
model_clade <- MCMCglmm(fixed  = formula_clade,
                        family = rep("gaussian", sum(select_traits)),
                        rcov   = clade_residuals,
                        data   = mcmcglmm_data,
                        nitt   = MCMCparams$nitt,
                        burnin = MCMCparams$burnin,
                        thin   = MCMCparams$thin)

## Phylogeny model
#GT: Now we will add phylogeny. To begin with we will not include the clade (gull, plover, sandpiper) effect and instead just partition covariances into phylogenetic and non-phylogenetic effects. At this point the model is becoming more complex and we need to specific a prior.
## Priors for the phylo model
#TODO: select the priors
nu <- 0.002
prior_phylo <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu)),
                    G = list(G1 = list(V = diag(sum(select_traits)), nu = nu)))
## The random terms for the phylo model
random_phylo <- ~ us(trait):animal
## The phylo model
#TODO: this is prone to crash due to random samples in the MCMC (I think it's rounding problems with the Cholesky decomposition)
model_phylo <- MCMCglmm(fixed = formula_noclade,
                        family = rep("gaussian", sum(select_traits)),
                        random = random_phylo,
                        rcov   = base_residuals,
                        prior  = prior_phylo,
                        pedigree = tree,
                        data   = mcmcglmm_data,
                        nitt   = MCMCparams$nitt,
                        burnin = MCMCparams$burnin,
                        thin   = MCMCparams$thin)

## Phylo and clade model
#GT: Now we will add phylogeny with the clade (gull, plover, sandpiper) effect. This model has a single overall phylogenetic effects and clade specific non-phylogenetic effects. Note the changes to the prior specification as well as the main model. This is the model that I would suggest using to test for changes in clade covariances relative to phylogeny to test which clades innovate/elaborate

#TODO: make the priors choice
prior_phylo_clade <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu),
                                   R2 = list(V = diag(sum(select_traits)), nu = nu)), 
                          G = list(G1 = list(V = diag(sum(select_traits)), nu = nu)))

## The phylo+clade model
model_phy_clade <- MCMCglmm(fixed  = formula_clade,
                            family = rep("gaussian", sum(select_traits)),
                            random = random_phylo,
                            rcov   = clade_residuals,
                            prior  = prior_phylo_clade,
                            pedigree = tree,
                            data   = mcmcglmm_data,
                            nitt   = MCMCparams$nitt,
                            burnin = MCMCparams$burnin,
                            thin   = MCMCparams$thin)

## Phylo_clade and clade
#GT: Now we will add phylogeny with phylogenetic and non-phylogenetic clade (gull, plover, sandpiper) effects. This model has a single overall phylogenetic effects and clade specific non-phylogenetic effects. Note the changes to the prior specification as well as the main model. This is the model that I would suggest using to test for changes in clade covariances relative to phylogeny to test which clades innovate/elaborate

## The phylo_clade random terms
random_phylo_clade <- ~ us(at.level(clade,1):trait):animal + us(at.level(clade,2):trait):animal

#TODO: make these priors automatic (again!)
prior_phy_cla_cla <- list(R = list(R1 = list(V = diag(sum(select_traits)), nu = nu),
                                   R2 = list(V = diag(sum(select_traits)), nu = nu)), 
                          G = list(G1 = list(V = diag(sum(select_traits)), nu = nu),
                                   G2 = list(V = diag(sum(select_traits)), nu = nu)))

## The phylo_clade + clade model
model_phy_clade <- MCMCglmm(fixed  = formula_clade,
                            family = rep("gaussian", sum(select_traits)),
                            random = random_phylo_clade,
                            rcov   = clade_residuals,
                            prior  = prior_phy_cla_cla,
                            pedigree = tree,
                            data   = mcmcglmm_data,
                            nitt   = MCMCparams$nitt,
                            burnin = MCMCparams$burnin,
                            thin   = MCMCparams$thin)
```
